### 2015-07-31 Harris_2015/manuscript-materials/draft.Rmd
more

\setlength{\parindent}{1cm}
\setlength{\parskip}{6pt}

@@

#### Conditional relationships and Markov networks.
Ecologists are often interested in learning about direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall")
relationships. The example with the shrubs and trees in Figure 1 shows how the
two measures can have opposite signs, and suggests that conditional
relationships are more closely related to the underlying ecological interactions.

Markov networks provide a method for translating back and forth between the
conditional relationships among species and the kinds of species assemblages
that these relationships produce. In this section, I show how a set of
conditional relationships can be used to determine the kinds of species
assemblages that can occur. Methods for estimating conditional relationships
from presence-absence data are discussed in the next section.

@@

\raggedright

@@

Of course, if all else is *not* equal (e.g. Figure 1,
where the presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from this
expectation. Predicting the marginal relationships between species from their
conditional interactions entails summing over the different possible assemblages
(Figure 2B), and the number of possible assemblages can be intractably large.
Numerous techniques are available for keeping the calculations feasible,
e.g. via analytic approximations [@lee_learning_2012] or Monte Carlo
sampling [@salakhutdinov_learning_2008], but they are beyond
the scope of this paper.  Alternatively, as noted below, some common linear
and generalized linear methods can also be used as computationally efficient
approximations to the full network.

@@

For the Markov networks, I used the rosalia package [@harris_rosalia_2015],
as described above. For the analyses with 20 species, I added a very weak
logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2 to
ensure that the model estimates were always finite. The bias introduced by this
prior should be small: the 95% credible interval on $\beta$ indicates that one
species' presence should not generally affect the odds of observing a different
species by more than a factor of 1500 (which is not much of a constraint). The
logistic distribution was chosen because it is convex and has a similar shape to
the Laplace distribution used in LASSO regularization (especially in the tails),
but it is differentiable everywhere and does not force any estimates to be
exactly zero.

I also evaluated five alternative methods from the existing literature to
calibrate the performance of the Markov network as a method for estimating
species interactions. The first two alternative interaction measures were the
sample correlations and the partial correlations between each pair of species'
data vectors on the landscape [@albrecht_spatial_2001; @faisal_inferring_2010].
Because partial correlations are undefined for landscapes with
perfectly-correlated species pairs, I used a regularized estimate based on
James-Stein shrinkage, as implemented in the corpcor package’s `pcor.shrink`
function with the default settings [@REF]. In the context of non-Gaussian
data, the partial correlation can be thought of as a computationally efficient
approximation to the full Markov network model [@loh_structure_2013]. This sort
of model is very common for drawing inferences about interaction networks in
bioinformatics contexts [@REF].

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit
regularized logistic regression models for each species, using
the other species on the landscape as predictors. [[Observational
presence-absence data does not have enough degrees of freedom to estimate this
many parameters, so I symmetrized the pairwise correlations by averaging the
coefficient for predicting species $i$’s status from species $j$ and vice
versa.]]

@@

The last two methods used the latent correlation matrix estimated by the
BayesComm package [@REF] to evaluate the claim from recent papers that the
correlation coefficients estimated by “joint species distribution models”
provide an accurate assessment of species’ pairwise interactions
[@pollock_understanding_2014; see also @harris_generating_2015].  I reported
both the average correlation between each pair of species’ latent variables
across 1000 Monte Carlo samples, as well as the average *partial* correlation.

@@

For the larger landscapes, I evaluated the relationship between each method's
estimates and the "true" interaction strengths among all of the species that
were observed to occur across the simulated landscapes. This kind of evaluation
is important because ecologists need a consistent way to know how strong species
interactions are regardless of which species were present in a particular data
set or how many observations were taken. Because the different methods mostly
describe species interactions on different scales (e.g. correlations versus $Z$
scores versus regression coefficients), I used linear regression through the
origin to rescale the different estimates produced by each method so that they
had a consistent interpretation. After rescaling each method’s estimates, I
calculated squared errors between the scaled interaction estimates and “true”
interaction values for each of the [[N]] species pairs across all the simulated
data sets. These squared errors determined the proportion of variance explained
for different combinations of method and landscape size (compared with a
baseline model that assumed all interaction strengths to be zero).

@@

#### Twenty species.
The accuracy of the evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3a). In general,
however, there was a consistent ordering.  The Markov network explained the
largest portion of the variance (34.6% overall), followed
by the generalized linear models (29.9%), partial correlations from the raw
presence-absence data (27.5%), and partial correlations from BayesComm, the joint
species distribution model (26.1).  The three methods that did not attempt to
control for indirect interactions all explained less than 20% of the variance.
Of these, the sample correlation matrix based on the raw data performed the best
(18.6%), followed by the null model (15.4) and BayesComm's correlation matrix
(11.6).

Although these last three methods had very different $R^2$ values, there was a
very close mapping between their estimates (after controlling for the size of
the simulated landscapes; Figure 3b).  This suggests that the correlation
matrices from joint species distribution models and especially the
effect sizes from null models contain essentially the same information that
could be obtained more easily by calculating correlation coefficients between
species' presence-absence vectors. This, in turn, demonstrates that these
methods do not effectively control for any indirect effects in the network.

@@

It’s important to note that none of these methods can identify the exact nature
of the pairwise interactions (e.g. which species in a positively-associated
  pair is facilitating the other), particularly when real pairs of species can
reciprocally influence one another in multiple ways simultaneously
[@bruno_inclusion_2003]; with compositional data, there is only enough
information to provide a single number describing each species pair.  
On the other hand, if ecologists are familiar enough with the natural history
of their study systems that they can augment the likelihood function with a
prior distribution or a suitable penalty, then this information could reduce the
effective degrees of freedom to estimate and real-world results might be even
better than those shown in Figure 3.

@@

Markov networks---particularly the Ising model for binary networks---have been
studied in statistical physics for nearly a century [@cipra_introduction_1987],
and the models' properties, capabilities, and limits are well-understood in a
huge range of applications. Modeling species interactions using the same
framework would thus allow ecologists to tap into an enormous set of existing
discoveries and techniques for dealing with indirect effects, stability, and
alternative stable states. Numerous other extensions are possible: Markov
networks can be fit with a mix of discrete and continuous variables, for example
[@lee_learning_2012]. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow the coefficient linking two species
in an interaction matrix to vary as a function of the abiotic environment or
of third-party species that tip the balance between facilitation and
exploitation [@bruno_inclusion_2003].

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about species
interactions. Null and neutral models can be very useful for clarifying our
thinking about the numerical consequences of species' richness and abundance
patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
null model must be interpreted with care [@roughgarden_competition_1983]. In
complex networks of ecological interactions (and even in small networks with
three species), it may simply not be possible to implicate individual species
pairs or specific ecological processes like competition by rejecting a
general-purpose null [@gotelli_empirical_2009]. Moreover, the null model’s
estimates were extremely similar to what could be obtained from a simple
correlation matrix, which raises questions about how much additional value the
null model adds. Estimating pairwise coefficients directly seems like a much
more promising approach: to the extent that the models' relative performance on
real data sets is similar to the range of results shown in Figure 3, scientists
in this field could [[easily double their explanatory power by switching from
null models to partial correlations or generalized linear models, or triple it
by switching to a Markov network]]. Regardless of the specific methods
ecologists ultimately choose, the most important consideration is clearly that
we need to control for indirect effects and estimate conditional relationships
between species, rather than marginal ones.

#### Acknowledgements:
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, N. J. Gotelli, and C. F. Dormann provided useful feedback on earlier
drafts of this work.

\setlength{\parindent}{0cm}
\setlength{\parskip}{4pt}

>>>>

\setlength{\parindent}{1cm}
\setlength{\parskip}{4pt}

@@

#### Markov networks.
Markov networks provide a method for translating back and forth between the
conditional relationships among species (Figure 1C) and the kinds of species
assemblages that these relationships produce. In this section, I show how a set
of conditional relationships can be used to determine the kinds of species
assemblages that can occur. Methods for estimating conditional relationships
from presence-absence data are discussed in the next section.

@@

\raggedright
\setlength{\parindent}{1cm}

@@

Of course, if all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then species'
marginal association rates can differ from this expectation. Predicting the
marginal relationships between species from their conditional interactions
entails summing over the different possible assemblages (Figure 2B), and the
number of possible assemblages can be intractably large, though some methods
beyond the scope of this paper can be employed to keep the calculations feasible
[@lee_learning_2012; @salakhutdinov_learning_2008]. Alternatively, as noted
below, some common linear and generalized linear methods can also be used as
computationally efficient approximations to the full network
[@lee_learning_2012; @loh_structure_2013].

@@

For the Markov networks, I used the rosalia package [@harris_rosalia_2015],
as described above. For the analyses with 20 species, I added a very weak
logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2 to
ensure that the model estimates were always finite. The bias introduced by this
prior should be small: the 95% credible interval on $\beta$ only requires that
one species' effect on the odds of observing a different species to be less than
a factor of 1500 (which is not much of a constraint). The logistic distribution
was chosen because it is convex and has a similar shape to the Laplace
distribution used in LASSO regularization (especially in the tails), but unlike
the Laplace distribution it is differentiable everywhere and does not force any
estimates to be exactly zero.

I also evaluated five alternative methods from the existing literature (plus a
novel variant of the fifth method) to calibrate the performance of the Markov
network as a method for estimating species interactions. The first two
alternative interaction measures were the sample correlations and the partial
correlations between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations
are undefined for landscapes with perfectly-correlated species pairs, I used a
regularized estimate based on James-Stein shrinkage, as implemented in the
corpcor package’s `pcor.shrink` function with the default settings
[@schafer_corpcor_2014]. In the context of non-Gaussian data, the partial
correlation can be thought of as a computationally efficient approximation to
the full Markov network model [@loh_structure_2013]. This sort of model is very
common for estimating relationships among genes and gene products [@REF].

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit
regularized logistic regression models [@gelman_weakly_2008] for each species,
using the other species on the landscape as predictors. [[Observational
presence-absence data does not have enough degrees of freedom to estimate this
many parameters, so I symmetrized the pairwise correlations by averaging the
coefficient for predicting species $i$’s status from species $j$ and vice
versa.]]

@@

The last two methods used the latent correlation matrix estimated by the
BayesComm package [@golding_bayescomm_2015] to evaluate the claim from recent
papers that the correlation coefficients estimated by "joint species
distribution models" provide an accurate assessment of species’ pairwise
interactions [@pollock_understanding_2014; see also @harris_generating_2015].  I
used both the average correlation between each pair of species’ latent
variables across 1000 Monte Carlo samples, as well as the average *partial*
correlation, as BayesComm's two estimators of species interactions.

@@

For the larger landscapes, I evaluated the relationship between each method's
estimates and the "true" interaction strengths among all of the species that
were observed to occur across the simulated landscapes. This kind of evaluation
is important because ecologists need a consistent way to know how strong species
interactions are regardless of which species were present in a particular data
set or how many observations were taken. Because the different methods mostly
describe species interactions on different scales (e.g. correlations versus $Z$
scores versus regression coefficients), I used linear regression through the
origin to rescale the different estimates produced by each method so that they
had a consistent interpretation. After rescaling each method’s estimates, I
calculated squared errors between the scaled interaction estimates and “true”
interaction values across all the simulated data sets. These squared errors
determined the proportion of variance explained for different combinations of
method and landscape size (compared with a baseline model that assumed all
interaction strengths to be zero).

@@

#### Twenty species.
The accuracy of the evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3a). In general,
however, there was a consistent ordering, with the four methods that controlled
for indirect effects of species interactions performing the best: the Markov
network explained the largest portion of the variance (34.6% overall), followed
by the generalized linear models (29.9%), partial correlations from the raw
presence-absence data (27.5%), and partial correlations from BayesComm, the
joint species distribution model (26.1%).  The benefit of choosing the full
Markov network over the other three methods was largest on the smaller
landscapes, which are also the ones that are most representative of typical
analyses in this field [[ref]].

The three methods that did not attempt to control for indirect interactions all
explained less than 20% of the variance. Of these, the sample correlation matrix
based on the raw data performed the best (18.6%), followed by the null model
(15.4%) and BayesComm's correlation matrix (11.6%). Although these last three
methods had different $R^2$ values, there was a close mapping between
their estimates (after controlling for the size of the simulated landscapes;
Figure 3b).  This suggests that the correlation matrices from joint species
distribution models and especially the effect sizes from null models only
contain noisy versions of the information that could be obtained more easily by
calculating correlation coefficients between species' presence-absence vectors.

@@

It’s important to note that none of these methods can identify the exact nature
of the pairwise interactions (e.g. which species in a positively-associated
pair is facilitating the other), particularly when real pairs of species can
reciprocally influence one another in multiple ways simultaneously
[@bruno_inclusion_2003]; with compositional data, there is only enough
information to provide a single number describing each species pair. To estimate
asymmetric interactions, such as commensalism or predation, ecologists would
need other kinds of data, as from time series, behavioral observations,
manipulative experiments, or natural history. These other sources of information
could also be used to augment the likelihood function with an informative prior
distribution, which could lead to better results on some real data sets than
was shown in Figure 3a.

@@

Markov networks---particularly the Ising model for binary networks---have been
studied in statistical physics for nearly a century [@cipra_introduction_1987],
and the models' properties, capabilities, and limits are well-understood in a
huge range of applications. Modeling species interactions using the same
framework would thus allow ecologists to tap into an enormous set of existing
discoveries and techniques for dealing with indirect effects, stability, and
alternative stable states. Numerous other extensions are possible: for example,
Markov networks can be fit with a mix of discrete and continuous variables,
and likely states of the interaction network can be modeled as a function of
the local abiotic environment [@lee_learning_2012]. There are even methods
[@whittam_species_1981; @tjelmeland_markov_1998] that would allow one species
to affect the sign or strength of the relationship between two other species,
tipping the balance between facilitation and exploitation
[@bruno_inclusion_2003].

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about species
interactions. Null and neutral models can be very useful for clarifying our
thinking about the numerical consequences of species' richness and abundance
patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
particular null model must be interpreted with care
[@roughgarden_competition_1983]. In complex ecological networks (and even in
small networks with three species), it may simply not be possible to implicate
individual species pairs or specific ecological processes like competition by
rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
test statistic is effectively just a correlation coefficient (Figure 3b).
Estimating pairwise coefficients directly seems like a much more promising
approach: to the extent that the models' relative performance on real data sets
is similar to the range of results shown in Figure 3a, scientists in this field
could often double their explanatory power by switching from null models to
Markov networks (or increase it by nearly as much with linear or generalized
linear approximations). Regardless of the specific methods ecologists ultimately
choose, the most important consideration is clearly that we need to control for
indirect effects and estimate conditional relationships between species, rather
than marginal ones.

#### Acknowledgements:
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, N. J. Gotelli, C. F. Dormann, and two anonymous reviewers also provided
useful feedback.

\setlength{\parindent}{0cm}
